---
title: "Problem Set 1"
subtitle: "API222: Big Data & Machine Learning"
author: "Brendan Hellweg"
date: "9/18/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(AER)
library(tidyverse)
library(class)
```

## Problem 1

#### Problem 1A
Regression, Inference. This is a regression problem because we are measuring continuous, quantifiable variables. It is an inference problem because we are seeking to understand current conditions to make a decision, not to predict future values.

#### Problem 1B
Classification, Prediction. This is a classification problem because the output of homelessness is a binary (yes, no). It is a prediction problem because we are estimating the likelihood of a future event.

#### Problem 1C
Regression, Inference. This is a regression problem because we are dealing with quantifiable attributes (including dummy variables with 0 or 1 values). It is inference because it is measuring relative risk in existing populations.

## Problem 2

#### Problem 2A
False. Low flexibility optimizes for low variance.

#### Problem 2B
True. Low flexibility models will have higher bias in these cases.

#### Problem 2C
False. The inflexible model will fit the data best.

#### Problem 2D
False. A model is parametric if f() can be finitely modeled, not if it has a decision rule. KNN models have decision rules.

## Problem 3
As flexibility increases, bias also increases due to overfitting, while variance decreases with the reduced MSE_TR. As flexibility decreases, bias also decreases because there is less overfitting, but variance will increase with the increased MSE_TR. As such, there is an inverse relationship between bias and variance, and the best models will optimize for the "sweet spot" in the middle that minimizes MSE_TE. 

\newpage

## Data Questions

```{r make_table,include=F}
data("MASchools") %>% 
  as.data.frame()

schools <- MASchools
```

### Problem 1
```{r prob1, echo=T}
nrow(schools)
```

### Problem 2
```{r prob2, echo=T}
ncol(schools)
```

### Problem 3
```{r prob3, echo=T}
sum(is.na(schools))
```

### Problem 4
The categorical variables are `district` and `municipality`.

### Problem 5
```{r prob5, echo=T}
mean(schools$score8,na.rm = T) %>% 
  round(.,2)
```

### Problem 6
```{r prob6, echo=T}
sd(schools$stratio) %>% 
  round(.,2)
```

### Setup for Regression & KNN Section
For the next section, I will filter the dataset to remove rows with NA values and create treatment and test bins.

```{r regsetup, echo=T}
schools2 <- schools %>% 
  drop_na() %>% 
  select(-c(1,2))

schools_tr <- schools2 %>% 
  head(.,n = (nrow(schools2)-35))

schools_te <- schools2 %>% 
  tail(.,n = 35)

set.seed(27)
```

### Problem 7
```{r prob7, echo=T}
school_lm <- lm(score8~.,schools_tr)

mean((schools_te$score8 - predict.lm(school_lm, schools_te)) ^ 2) %>% 
  round(.,2)
```

### Problem 8
```{r prob8, echo=T}
school_lm2 <- lm(score8~stratio+lunch+expreg,schools_tr)

mean((schools_te$score8 - predict.lm(school_lm2, schools_te)) ^ 2) %>% 
  round(.,2)
```

```{r prob8a, echo=T}
print(school_lm)
print(school_lm2)
```

The coefficient with the largest change in effect is `stratio`, which in the first model is associated with a moderate decline in 8th grade test performance, while in the second model it is associated with a moderate improvement in 8th grade test performance. In comparison, an increase in `lunch` continues to be associated with a decline in `score8`, though the coefficient roughly doubles in magnitude. The smallest coefficient change is for `expreg`, which changes from slightly negative to slightly positive.

In the first model, `stratio` has a coefficient of `-0.54`, while in the second model its coefficient is `0.49`.

One reason this change might occur is interaction effects missing from the all-variables linear model. When a model contains too many variables and insufficient flexibility, its bias may produce models with a high degree of error. Second, the many-variable model will have less interpretability, so these associations are harder to discuss.

### Problem 9
```{r prob9, echo=T}
set.seed(27)
pr <- knn(schools_tr,schools_te,schools_tr$score8,k = 2)
tb <- as.data.frame(pr) %>% 
  mutate(pr = as.numeric(as.character(pr))) %>% 
  mutate(scores = schools_te$score8) %>% 
  mutate(sqerror = as.numeric(as.character((.$pr-.$scores)^2))) 
 
mean(as.numeric(as.character(tb$sqerror))) %>% 
  round(.,2)
```

### Problem 10
```{r prob10, echo=T}
set.seed(27)
pr <- knn(schools_tr,schools_te,schools_tr$score8,k = 10)
tb <- as.data.frame(pr) %>% 
  mutate(pr = as.numeric(as.character(pr))) %>% 
  mutate(scores = schools_te$score8) %>% 
  mutate(sqerror = as.numeric(as.character((.$pr-.$scores)^2))) 
 
mean(as.numeric(as.character(tb$sqerror))) %>% 
  round(.,2)
```